\color{gray}

\setcounter{chapter}{-1}
\chapter{General Course Information}
\label{ch:general_info}

\section{Course Website}

For updates and detailed syllabus, refer to the official course website:
\begin{center}
    \url{https://www.unibo.it/en/study/course-units-transferable-skills-moocs/course-unit-catalogue/course-unit/2025/535171}
\end{center}

\section{Learning Outcomes}

Upon completion of this course, the student will be able to:

\begin{itemize}
    \item Describe the fundamental issues and principal applications of image processing.
    \item Demonstrate a solid understanding of state-of-the-art image processing methods.
    \item Identify, analyse, and address image processing problems by recognising and employing (or proposing) effective solutions.
    \item Design and implement practical solutions for a range of common image processing challenges.
\end{itemize}

Furthermore, the student will acquire fundamental knowledge regarding the main algorithms for:

\begin{itemize}
    \item Filtering in both spatial and frequency domains;
    \item Image segmentation;
    \item Object detection and recognition.
\end{itemize}

\section{Course Contents}

The course covers the following topics:

\begin{itemize}
    \item \textbf{Visual perception:} Mechanics of human vision and interpretation.
    \item \textbf{Image formation:} Geometry and radiometry.
    \item \textbf{Digitisation:} Transition from analogue to digital images.
    \item \textbf{The digital image chain:} Acquisition, visualisation, and processing.
    \item \textbf{Image processing techniques:}
    \begin{enumerate}
        \item Geometric operations;
        \item Point operations;
        \item Local operations;
        \item Image segmentation;
        \item Fourier analysis.
    \end{enumerate}
    \item \textbf{Advanced extraction:} Item segmentation and feature extraction.
    \item \textbf{Laboratory sessions:} Practical application using ImageJ.
\end{itemize}

\section{Assessment Methods}

During the final weeks of the course, students will address a real-world problem through a group project. The assessment involves:

\begin{itemize}
    \item \textbf{Group Project:} Work performed in small groups (4 or 5 students).
    \item \textbf{Literature Review:} Reading scientific papers and conducting bibliography/resource searches.
    \item \textbf{Implementation:} Understanding and implementing specific algorithms.
    \item \textbf{Presentation:} Delivery of both a written report and an oral presentation detailing the project and its outcomes.
\end{itemize}

\paragraph{Final Examination}
The final exam consists of a discussion regarding the group project (distinct from the initial presentation) followed by an oral interview covering the course topics.

% Box modificato per usare la versione leggera definita in setup.tex
\paragraph{Important Note}
    The material available on the course website is not exhaustive. In certain cases, consulting the recommended textbooks and further reading material is necessary.